---
title: "Linear Regression (using BAS)"
output: html_notebook
---


##  Warm up: simple linear regression.

For our example we use a dataset contained in the library BAS. We first uplaod the library and see the row data 
"bodyfat". 

```{r}
library(BAS)
data(bodyfat)
bodyfat$Density=c() # revmove this variable 
summary(bodyfat)
#bodyfat$Bodyfat=(bodyfat$Bodyfat-mean(bodyfat$Bodyfat))/sd(bodyfat$Bodyfat)
#help(bodyfat)
```

This data frame includes 252 observations of men’s body fat and other measurements, such as waist circumference (Abdomen). See help(bodyfat) in R. 


As warming up exercise we start by a simple linear regression:  
$$
y_i = \beta_1 x_{i1} + \beta_0 +\epsilon_i \quad \epsilon_i \stackrel{iid}{\sim} N(0,1)
$$


We use as $y_i$ the variable Bodyfat and as covariate $x_{i1}$ the variable Abdomen. We start by fitting  a frequentistic Linear Model (simple regression)

```{r}
bodyfat.lm = lm(Bodyfat ~ Abdomen, data = bodyfat)
summary(bodyfat.lm)
beta = coef(bodyfat.lm)
```

Let us plot the regression line. 

```{r}
plot(bodyfat$Abdomen,bodyfat$Bodyfat,xlab="abdomen circumference (cm)",ylab="body fat")
abline(beta,col='red')
```



We can easily obtain residuals,  compute the MS,  plot the residuals and make the QQ-plot.

REMEMBER: all this stuff if NOT Bayesian!!! We do this for comparison!! (bad experience with previous students!)
 
```{r}
resid = residuals(bodyfat.lm)
n = length(resid)
MSE = 1/ (n - 2) * sum((resid ^ 2))
qqnorm(resid/sd(resid)) 
qqline(resid/sd(resid))
plot(resid)
abline(h=0,col="red")
```
```{r}
MSE
```



# Linear regression

In a normal linear regression problem, the likelihood is

$$
y=(y_1,\dots,y_n)  \sim \mathscr{N}_{n}(\beta_0 \mathbf{1} + \beta X, \sigma^2 I_n) 
\quad \beta=(\beta_1,\dots,\beta_k)
$$

where $\mathbf{1}$ is a vector of ones of length $n$ and $I_n$ is the identity matrix.

Again as warm up exercise, let us  fit the  full model above  in a frequentistic way (MLE)
using R. 

```{r}
fulllm=lm(Bodyfat ~ ., data = bodyfat)
summary(fulllm)
```



# Bayesian regression using BAS


We now use the BAS package to fit a Bayesian Linear Regression model. BAS is mainly designed to fit a hierarchy of nested models (including and excluding the various covariates) in order to perform model 
selection and model averaging. Here we  use BAS just to fit one single (given) model.
More will be done in the LAB sessions with the TA. 

We first fit the model using a g-prior (Zellner) with constant=20.  This is done by specifying 
the string "modelprior = Bernoulli(1), include.always = ~ ., n.models = 1". Model prior is the prior over the various models, in this case using a Ber(1) we include with probability 1 all the covariates in the prior (and hence also in the posterior). This corresponds to a prior which gives mass 1 to the complete model. 
The command n.models =1 specifies that we take into consideration only the first model (in this case the full model, due to the form of the prior). 
Finally "include.always = ~ ."  specifies the fact that in the model we include the intercept $\beta_0$. 

Omitting the prior description on $\beta_0$, recall that the (centered) Zellner prior is 

$$
\begin{split}
& \beta=(\beta_1,\dots,\beta_k) \\
\beta|\sigma^2,X &\sim \mathscr{N}_{k}(0,\alpha \sigma^2(X^t X)^{-1})\\
\sigma^2|X  & \sim \pi(\sigma^{2})=\sigma^{-2}  \qquad   \alpha>0.
\end{split}
$$


In the book $\alpha$ is $c$. 



We take $\alpha=30$ and fit the model with BAS. 

```{r}
alphapar=30
cog.bas = bas.lm(Bodyfat ~ ., data =  bodyfat, prior="g-prior",alpha=alphapar, modelprior = Bernoulli(1), include.always = ~ ., n.models = 1)
beta = coef(cog.bas)
beta
```


We plot the posterior densities of the $\beta$s. 

```{r}
plot(beta, subset = 1:10, ask = F)
```

To obtain credible intervals for coefficients, BAS includes a confint method to create Highest Posterior (5%) Bodyfat intervals for  the summaries from coef.



```{r}
confint(beta)
plot(confint(beta),main=paste("g-prior alpha=",alphapar))

```


We now choose a suitable mxitrue of Zellner prior, more specifically the Zellner-Siow prior. A ZS prior is simply a mixture of Zellner prior, i.e. 
$$
\begin{split}
& \beta=(\beta_1,\dots,\beta_k) \\
\beta|\sigma^2,X &\sim \mathscr{N}_{k}(0,\alpha \sigma^2(X^t X)^{-1})\\
\sigma^2|X  & \sim \pi(\sigma^{2})=\sigma^{-2}  \\
& 1/\alpha \sim \pi_0 = Gamma(1/2,n/2)
\end{split}
$$

Note that choosing other form for $\pi_0$ one obtains other Zellner type priors, some of them are implemented in BAS. 



```{r}
cog.basZS = bas.lm(Bodyfat ~ ., data =  bodyfat, prior="JZS", modelprior = Bernoulli(1), include.always = ~ ., n.models = 1)
#plot(cog.basZS, which=4)
betaZS = coef(cog.basZS)
betaZS
plot(betaZS,subset = 2:7, ask = F)
```

Let us compare the $\beta$s estimates under the two different priors


```{r}
par(mfrow=c(2,2))
plot(confint(beta),main=paste("g-prior alpha=",alphapar))
confint(betaZS)
plot(confint(betaZS),main="ZS-prior ")
```


BAS has methods defined to return fitted values using the observed design matrix and predictions at either the observed data or potentially new values, as with lm.


```{r}
n=40
nend=length(bodyfat[,1])
newdata<-bodyfat[1:n,]
datalearning<-bodyfat[seq(n+1,nend),]
#dim(datalearning)
rownames(datalearning)=seq(1:length(datalearning[,1]))

cog.basZS2 = bas.lm(Bodyfat ~ ., data =  datalearning, prior="JZS", modelprior = Bernoulli(1), include.always = ~ ., n.models = 1)
betaZS2 = coef(cog.basZS2)
#plot(confint(betaZS2),main="ZS-prior ")



fitted<-predict(cog.basZS2, estimator = "BMA")
prednew <- predict(cog.basZS2,newdata=newdata, estimator = "BMA")
```

Below we plot true values vs fitted and predicted. Black points are 
true vs fitted in the learning dataset and red points are true vs predicted in the additional part of the data set (not used for fitting the model).


```{r}
plot(fitted$Ypred[1:length(fitted$Ypred)],datalearning$Bodyfat[1:length(fitted$Ypred)],
  pch = 16,
  xlab = expression(hat(mu[i])), ylab = 'Y',type="p")


points(prednew$Ypred, newdata$Bodyfat,
  pch = 16,
  col="red",type="p"
)
abline(0, 1)


#prednew$Ypred 
#newdata$Bodyfat


BPM <- predict(cog.basZS2, estimator = "BPM", newdata=newdata,se.fit = TRUE)
conf.fit <- confint(BPM, parm = "mean")
conf.pred <- confint(BPM, parm = "pred")
plot(conf.pred, main="Out of sample: pred. (black) vs true (red)")
points(seq(1:n),newdata$Bodyfat,col="red")

```




For much more see https://cran.r-project.org/web/packages/BAS/vignettes/BAS-vignette.html

see also 
https://www.sciencedirect.com/science/article/pii/S030440761200156X
